---
title: "Practical assignments"
author: "Vitória Bizão Murakami"
format:
  html:
    embed-resources: true  # Imagens e scripts dentro do próprio HTML 
    toc: true              # Sumário?
    toc-location: left     # Local do sumário
    code-overflow: wrap    # Não deixa o código passar a margem
    code-fold: true        # Esconde o código
    fig-align: center      # Centraliza no HTML
    fig-width: 10          # Aumenta a largura padrão no HTML
    grid:
      sidebar-width: 250px # Tamanho da barra lateral
      body-width: 1000px   # Tamanho do corpo do arquivo
      margin-width: 250px  # Tamanho da margem
      gutter-width: 1.5rem # Espaçamento entre os elementos acima
  pdf:
    fig-align: center      # Centraliza as imagens
    fig-pos: 'H'           # Força a figura a ficar "Aqui"
    include-in-header:     # Não deixa o código passar a margem
      text: |
        \usepackage{fvextra}
        \usepackage{float}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
execute:
  fig-width: 8             # Largura padrão global
  fig-height: 6            # Altura padrão global
---

# Using your USP number as seed, simulate phenotypic values (don’t forget, P=G+E) using the normal distribution (rnorm). Then, compute the mean and the variance of P, G and E, and build a histogram, a density plot, a box plot and a scatter plot depicting the relationship between P and G. Do it for three different sample sizes: 10, 100 and 1000.

```{r}
#| message: false
#| warning: false
pkg = c("tidyverse", "readxl", "gridExtra", "ggbeeswarm", "grid", "ggtext", "cowplot")
sapply(pkg, require, character.only=TRUE)

set.seed(11480230)

## Amostra = 10
Genotype <- rnorm(n = 10, mean = 0, sd = 100)
Environment <- rnorm(n = 10, mean = 0, sd = 10)
Phenotype <- Genotype + Environment

df <- data.frame(Genotype, Environment, Phenotype); df

### Histogram
p1 <- df |> 
  # Transforming to longer format
  pivot_longer(cols = c(Genotype, Environment, Phenotype), 
               names_to = "origem", 
               values_to = "Caracter") |> 
  # Creating the graphic
  ggplot(aes(x = Caracter, fill = origem)) + 
  geom_histogram(bins = 20, color = 'white', show.legend = FALSE) + 
  
  # Faceting
  facet_wrap(~ origem, ncol = 1, scales = "free_y") + 
  
  # Design
  scale_fill_manual(values = c("Genotype" = "#cb73fa",
                               "Environment" = "purple",
                               "Phenotype" = "purple4")) +
  labs(x = "Caracter",
       y = "Count",
       title = "G, P and E - Histogram") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14, face = "bold"),
        strip.text = element_text(face = "bold")) 

### Density
p2 <- df |> 
  # Transforming to longer format
  pivot_longer(cols = c(Genotype, Environment, Phenotype), 
               names_to = "origem", 
               values_to = "Caracter") |> 
  
  # Creating graphic
  ggplot(aes(x = Caracter, color = origem)) + 
  geom_density(linewidth = 1) + 
  
  # Design
  scale_color_manual(values = c("Genotype" = "#cb73fa",
                                "Environment" = "purple",
                                "Phenotype" = "purple4")) +
  labs(x = "Caracter", 
       y = "Count",
       title = "G, E and P - Density",
       color = "Origem") + 
  theme_minimal() + 
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        plot.title = element_text(face = "bold", size = 14))

### Boxplot
p3 <- df |> 
  # 1. Formato longo para g, e, p
  pivot_longer(cols = c(Genotype, Environment, Phenotype),
               names_to = "origem",
               values_to = "Caracter") |> 
  
  # 2. Início do Plot
  ggplot(aes(x = Caracter, y = 1, fill = origem)) + 
  
  # Boxplot horizontal
  geom_boxplot(color = "black", alpha = 0.8, show.legend = FALSE) + 
  
  # 3. Substituir o annotate por stat_summary para calcular a média por facet
  stat_summary(fun = mean, geom = "point", shape = 17, size = 4, color = "purple4") +
  
  # 4. Facets: uma linha para cada origem
  facet_wrap(~ origem, ncol = 1, strip.position = "left") + 
  
  # Estética
  scale_fill_manual(values = c("Genotype" = "#cb73fa",
                               "Environment" = "purple",
                               "Phenotype" = "purple4")) +
  labs(x = "Caracter",
       y = NULL,
       title = "G, E, and P - Boxplot") +
  theme_minimal() + 
  theme(legend.position = "none",
        legend.title = element_blank(),
        plot.title = element_text(size = 14, face = "bold"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text = element_text(face = "bold"))

### Scatterplot GxP
p4 <- df |> ggplot(aes(x = Genotype, y = Phenotype)) + 
  geom_jitter(alpha = .3, color = "purple") + 
  labs(x = "Genotype",
       y = "Phenotype",
       title = "Genotype vs. Phenotype - Scatterplot") + 
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# Grid
grid.arrange(p1, p2, p3, p4,
             ncol = 2,
             top = grid::textGrob("Plots for n =",
                                  gp = grid::gpar(fontsize = 14, fontface = "bold")))

# Transforming in a function
samples_dif <- function(n1, n2, n3) {
  set.seed(11480230)   # Setting seed
  sample_sizes <- c(n1, n2, n3) # Receive the sample sizes 
  results <- list() # Store results
  plots <- list() # Store plots
  
  for (i in seq_along(sample_sizes)) {
    n <- sample_sizes[i]
    
    # Generating the data
    Genotype <- rnorm(n = n, mean = 0, sd = 10)
    Environment <- rnorm(n = n, mean = 0, sd = 20)
    Phenotype <- Genotype + Environment
    df <- data.frame(Genotype, Environment, Phenotype)
    results[[i]] <- df
    
    # Plots
    ### Histogram
    p1 <- df |> 
      pivot_longer(cols = c(Genotype, Environment, Phenotype), 
                   names_to = "origem", 
                   values_to = "Caracter") |> 
      ggplot(aes(x = Caracter, fill = origem)) + 
      geom_histogram(bins = 20, color = 'white', show.legend = FALSE) + 
      facet_wrap(~ origem, ncol = 1, scales = "free_y") + 
      scale_fill_manual(values = c("Genotype" = "#cb73fa",
                                   "Environment" = "purple",
                                   "Phenotype" = "purple4")) +
      labs(x = "Caracter",
           y = "Count",
           title = "G, P and E - Histogram") + 
      theme_minimal() + 
      theme(plot.title = element_text(size = 14, face = "bold"),
            strip.text = element_text(face = "bold"))
    
    ### Density
    p2 <- df |> 
      pivot_longer(cols = c(Genotype, Environment, Phenotype), 
                   names_to = "origem", 
                   values_to = "Caracter") |> 
      ggplot(aes(x = Caracter, color = origem)) + 
      geom_density(linewidth = 1) + 
      scale_color_manual(values = c("Genotype" = "#cb73fa",
                                    "Environment" = "purple",
                                    "Phenotype" = "purple4")) +
      labs(x = "Caracter", 
           y = "Count",
           title = "G, E and P - Density",
           color = "Origem") + 
      theme_minimal() + 
      theme(legend.position = "bottom",
            legend.title = element_blank(),
            plot.title = element_text(face = "bold", size = 14))
    
    ### Boxplot
    p3 <- df |> 
      pivot_longer(cols = c(Genotype, Environment, Phenotype),
                   names_to = "origem",
                   values_to = "Caracter") |> 
      ggplot(aes(x = Caracter, y = 1, fill = origem)) + 
      geom_boxplot(color = "black", alpha = 0.8, show.legend = FALSE) + 
      stat_summary(fun = mean, geom = "point", shape = 17, size = 4, color = "purple4") +
      facet_wrap(~ origem, ncol = 1, strip.position = "left") + 
      scale_fill_manual(values = c("Genotype" = "#cb73fa",
                                   "Environment" = "purple",
                                   "Phenotype" = "purple4")) +
      labs(x = "Caracter",
           y = "",
           title = "G, E, and P - Boxplot") +
      theme_minimal() + 
      theme(legend.position = "none",
            legend.title = element_blank(),
            plot.title = element_text(size = 14, face = "bold"),
            axis.text.y = element_blank(),
            axis.ticks.y = element_blank(),
            strip.text = element_text(face = "bold"))
    
    ### Scatterplot GxP
    p4 <- df |> ggplot(aes(x = Genotype, y = Phenotype)) + 
      geom_jitter(alpha = .3, color = "purple") + 
      labs(x = "Genotype",
           y = "Phenotype",
           title = "Genotype vs. Phenotype - Scatterplot") + 
      theme_minimal() +
      theme(plot.title = element_text(size = 14, face = "bold"))
    
    # 2. Criar o Grid para esta amostra n
    grid_n <- grid.arrange(p1, p2, p3, p4,
                           ncol = 2,
                           top = grid::textGrob(paste("Plots for n =", n),
                                                gp = grid::gpar(fontsize = 14, fontface = "bold")))
    
    plots[[i]] <- grid_n
  }
  
  # 3. Tabela Resumo (sample_summary)
  sample_summary <- data.frame(
    "Sample_Size" = sample_sizes,
    "Genotypic_Mean" = sapply(results, function(x) mean(x$Genotype)),
    "Genotypic_Var" = sapply(results, function(x) var(x$Genotype)),
    "Env_Mean" = sapply(results, function(x) mean(x$Environment)),
    "Env_Var" = sapply(results, function(x) var(x$Environment)),
    "Pheno_Mean" = sapply(results, function(x) mean(x$Phenotype)),
    "Pheno_Var" = sapply(results, function(x) var(x$Phenotype))
  )
  
  return(list(table = sample_summary, plots = plots))
}

# Testing function
output <- samples_dif(10, 100, 1000)

# Results summary
output$table

# Plots n = 10
plot(output$plots[[1]])

# Plots n = 100
plot(output$plots[[2]])

# Plots n = 1000
plot(output$plots[[3]])
```

# Choose a random trial and perform exploratory data analysis with three traits: GY, PH and NDM. Then, compute the correlation between these traits and build pairwise scatter plots.

```{r}
#| message: false
#| warning: false
df = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';')

# selecting the traits in trial 2
df <- df |> 
  filter(rep == 2) |> 
  select(gen, GY, PH_R8, NDM, OIL) |> 
  mutate(gen = as.factor(gen))

# EDA 
# Histograms
h_GY <- df |> ggplot(aes(x = GY)) +
    geom_histogram(aes(y = ..density..), fill = "purple4", alpha = 0.75, binwidth = 70) +
    geom_density(fill = "purple", alpha = 0.2, color = "black") +
    labs(title = "GY",
         x = "GY",
         y = "Frequency") +
    theme_minimal() + 
    theme(plot.title = element_text(face = "bold", size = 14))

h_PH <- df |> ggplot(aes(x = PH_R8)) +
  geom_histogram(aes(y = ..density..), fill = "purple4", alpha = 0.75, binwidth = 10) +
  geom_density(fill = "purple", alpha = 0.2, color = "black") +
  labs(title = "PH_R8",
       x = "PH_R8",
       y = "Frequency") +
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

h_NDM <- df |> ggplot(aes(x = NDM)) +
  geom_histogram(aes(y = ..density..), fill = "purple4", alpha = 0.75, binwidth = 8) +
  geom_density(fill = "purple", alpha = 0.2, color = "black") +
  labs(title = "NDM",
       x = "NDM",
       y = "Frequency") +
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

h_OIL <- df |> ggplot(aes(x = OIL)) +
  geom_histogram(aes(y = ..density..), fill = "purple4", alpha = 0.75, binwidth = 0.5) +
  geom_density(fill = "purple", alpha = 0.2, color = "black") +
  labs(title = "OIL",
       x = "OIL",
       y = "Frequency") +
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

grid.arrange(h_GY, h_PH, h_NDM, h_OIL, ncol = 2)

# Loop - Scatterplots
vars <- c("GY", "PH_R8", "NDM", "OIL")
pares <- combn(vars, 2, simplify = FALSE)
resultados <- list()

for (par in pares) {
  x_var <- par[1]
  y_var <- par[2]
  nome_plot <- paste0("scatter_", x_var, "_vs_", y_var)
  
  message("Gerando Scatter: ", x_var, " vs ", y_var)
  
  resultados[[nome_plot]] <- ggplot(df, aes(x = .data[[x_var]], y = .data[[y_var]])) + 
    geom_jitter(color = "gray20", size = 0.75, alpha = 0.2) + 
    geom_smooth(method = "auto", se = TRUE, color = "#cb73fa") +
    labs(title = paste(x_var, "vs", y_var),
         x = x_var,
         y = y_var) + 
    theme_classic() + 
    theme(axis.title = element_text(size = 12, face = "bold"),
          plot.title = element_text(size = 13, face = "bold"))
}

scatters_only <- resultados[grep("scatter", names(resultados))]
grid.arrange(grobs = scatters_only, ncol = 3)

# Correlation
df[,-1] |> GGally::ggpairs()
```
