---
title: "Genética-Estatística I"
author: "Vitória Bizão Murakami"
format:
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center      # Centraliza no HTML
    fig-width: 10          # Aumenta a largura padrão no HTML
    grid:
      sidebar-width: 250px
      body-width: 1000px   # Já está bom, permite figuras largas
      margin-width: 250px
      gutter-width: 1.5rem
  pdf:
    fig-align: center      # Centraliza no PDF
    fig-pos: 'H'           # Força a figura a ficar "Aqui"
    include-in-header:
      text: |
        \usepackage{fvextra}
        \usepackage{float}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
execute:
  fig-width: 8             # Largura padrão global (em polegadas)
  fig-height: 6
---

# Task 2
Read Chapter 10 of R for Data Science and solve all exercises. Build a report using Quarto or R Markdown and render it in pdf or html. 

```{r}
#| message: false
#| warning: false
#| include: false
# Limpando o ambiente:
rm(list = ls())

# Carregando vários pacotes a serem usados na sessão simultaneamente:
pkg = c("tidyverse", "dplyr", "tidyr", "ggplot2", "gridExtra", "nycflights13", "ggbeeswarm")
sapply(pkg, require, character.only=TRUE)
```

# 10.3.3 Exercises
## Exercise 1

Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

```{r}
#| message: false
#| warning: false
# X
# Loading dataset
df <- diamonds

x <- df |>  
  ggplot(aes(x = x)) +
  geom_freqpoly(binwidth = 0.05, size = .8, color = "#cb73fa") +
  scale_x_continuous(breaks = seq(0,10,1), limits = c(0,10)) +
  labs(x = "Length",
       y = "Count", 
       title = "Frequency of diamond length (in mm)",
       tag = "X") +
  theme_classic() +
  theme(axis.title = element_blank(),
        title = element_blank(),
        plot.tag = element_text(size = 9))
# Y
y <- df |> 
  filter(y > 3.5 & y < 15) %>% 
  ggplot(aes(x = y)) +
  geom_freqpoly(binwidth = 0.05, linewidth = .8, color = "purple") +
  scale_x_continuous(breaks = seq(0,10,1), limits = c(0,10)) +
  labs(x = "Width",
       y = "Count", 
       title = "Frequency of diamond length (in mm)",
       tag = "Y") +
  theme_classic() +
  theme(axis.title = element_blank(),
        title = element_blank(),
        plot.tag = element_text(size = 9))
# Z 
z <- df |> 
  ggplot(aes(x = z)) +
  geom_freqpoly(binwidth = 0.05, size = .8, color = "purple4") +
  scale_x_continuous(breaks = seq(0,10,1), limits = c(0,10)) +
  labs(x = "Depth",
       y = "Count", 
       title = "Frequency of diamond length (in mm)",
       tag = "Z") +
  theme_classic() +
  theme(axis.title = element_blank(),
        title = element_blank(),
        plot.tag = element_text(size = 9))

xyz <- grid.arrange(x, y, z, 
                    ncol = 1,
                    top = grid::textGrob("Diamond dimensions (in mm) - frequency plots",
                                         gp = grid::gpar(fontsize = 12, fontface = "bold")),
                    left = grid::textGrob("Count",
                                          rot = 90,
                                          gp = grid::gpar(fontsize = 10, fontface = "bold")),
                    bottom = grid::textGrob("Millimeters (mm)",
                                            gp = grid::gpar(fontsize = 10, fontface = "bold")))

plot(xyz)
```

## Exercise 2
Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)
```{r}
#| message: false
#| warning: false
df |>  
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 100,
                 color = "black",
                 fill = "#cb73fa") +
  geom_density(aes(y = after_stat(count * 100)), 
               color = "black",
               linewidth = 0.75) +
  labs(x = "Price (in USD)",
       y = "Frequency",
       title = "Diamond price") +
  scale_x_continuous(breaks = seq(0,20000,2500)) +
  scale_y_continuous(breaks = seq(0,5000,500)) +
  theme_minimal() + 
  theme(axis.title = element_text(size = 10, face = "bold"),
        plot.title = element_text(size = 12, face = "bold"))
```

## Exercise 3
How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?

Primeiro vou visualizar a distribuição dos quilates.
```{r}
#| message: false
#| warning: false
df |>  
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.03,
                 color = "black",
                 fill = "#cb73fa") +
  xlab("Weight of the diamond") +
  ylab("Frequency") +
  ggtitle("Diamond carat") +
  theme_minimal() + 
  theme(axis.title = element_text(size = 10, face = "bold"),
        plot.title = element_text(size = 12, face = "bold"))
```

Contagem de NAs:
```{r}
#| message: false
#| warning: false
tb <- bind_rows(df |> filter(carat == 1.00) |>
                  count() |>
                  mutate(carat = 1.00),
                df |> filter(carat == 0.99) |>
                  count() |>
                  mutate(carat = 0.99)) |>
  select(carat, n) 

knitr::kable(tb, 
      col.names = c("Carat", "Number of NAs"),
      align = "cc",
      caption = "NA Frequency: 1.00 vs 0.99 Carat")
```

```{r}
#| message: false
#| warning: false
## Loop
# Selecting variables
df_rn <- df |> 
  rename("Total Depth" = depth, "Table" = table, "Price" = price, "Length" = x, "Width" = y, "Depth" = z)

y_vars <- df_rn |> 
  select(where(is.numeric)) |> 
  names() |> 
  setdiff("carat")

# Null list to store data
plot_list <- list()

# Runing loop
for (y_var in y_vars) {
  plot_list[[y_var]] <- ggplot(df_rn, aes(x = carat, y = .data[[y_var]])) + 
    geom_jitter(color = "gray20", size = 0.75, alpha = 0.1) + 
    geom_smooth(method = "auto", se = TRUE, color = "#cb73fa") +
    labs(y = y_var) + 
    theme_classic() + 
    theme(axis.title = element_text(size = 10, face = "bold"),
          axis.title.x = element_blank())
}

# Grid with all graphics
do.call(grid.arrange, c(plot_list, 
                        list(ncol = 3, 
                             top = grid::textGrob("Diamond carat VS other variables", 
                                                  gp = grid::gpar(fontsize = 12, fontface = "bold")),
                             bottom = grid::textGrob("Carat",
                                            gp = grid::gpar(fontsize = 10, fontface = "bold")))))
```

## Exercise 4
Compare and contrast coord_cartesian() vs. xlim() or ylim() when zooming in on a histogram. What happens if you leave binwidth unset? What happens if you try and zoom so only half a bar shows?
```{r}
#| message: false
#| warning: false

# Binwidth unset
p1 <- ggplot(df, aes(x = y)) + 
  geom_histogram(color = "black",
                 fill = "#cb73fa") +
  coord_cartesian(ylim = c(0, 30000)) +
  labs(title = "Unset") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        title = element_text(size = 8))

# Setting binwidth
p2 <- ggplot(df, aes(x = y)) + 
  geom_histogram(color = "black",
                 fill = "#cb73fa",
                 binwidth = .5) +
  coord_cartesian(ylim = c(0, 30000)) +
  labs(title = "binwidth = 0.5") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        title = element_text(size = 8))

# Setting binwidth
p3 <- ggplot(df, aes(x = y)) + 
  geom_histogram(color = "black",
                 fill = "#cb73fa",
                 binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 2.5),
                  xlim = c(10, 40)) +
  labs(title = "bin, y and xlim set") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        title = element_text(size = 8))

# Binwidth unset
p4 <- ggplot(df, aes(x = y)) + 
  geom_histogram(color = "black",
                 fill = "#cb73fa") +
  coord_cartesian(xlim = c(0, 15)) +
  labs(title = "xlim = c(0, 15)") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        title = element_text(size = 8))

grid.arrange(p1, p2, p3, p4,
             ncol = 2,
             top = grid::textGrob("Testing coord_cartesian() Vs. xlim() and ylim()",
                                   gp = grid::gpar(fontsize = 12, fontface = "bold")),
             left = grid::textGrob("Count",
                                   rot = 90,
                                   gp = grid::gpar(fontsize = 10, fontface = "bold")),
             bottom = grid::textGrob("Width (in mm)",
                                   gp = grid::gpar(fontsize = 10, fontface = "bold")))
```

# 10.4.1 Exercises
## Exercise 1 
What happens to missing values in a histogram? What happens to missing values in a bar chart? Why is there a difference in how missing values are handled in histograms and bar charts?
```{r}
#| message: false
#| warning: false
df_na <- diamonds |> 
  mutate(y = if_else(y < 3 | y > 20, NA, y))

sum(is.na(df_na$y))

df_na |> ggplot(aes(x = y)) +
  geom_histogram(binwidth = 0.2,
                 color = "black",
                 fill = "#cb73fa") +
  labs(x = "Width",
       y = "Frequency",
       title = "Diamond width in mm") +
  scale_x_continuous(breaks = seq(0,10,1)) +
  scale_y_continuous(breaks = seq(0,8000,1000)) +
  theme_minimal() + 
  theme(axis.title = element_text(size = 10, face = "bold"),
        plot.title = element_text(size = 12, face = "bold"))
```

## Exercise 2
hat does na.rm = TRUE do in mean() and sum()?
```{r}
#| message: false
#| warning: false
knitr::kable(data.frame(`Metric` = c("Sum", "Mean"),
                        `Without na.rm` = c(sum(df_na$y), mean(df_na$y)),
                        `With na.rm = T` = c(round(sum(df_na$y, na.rm = TRUE),2),
                                             round(mean(df_na$y, na.rm = TRUE),2))),
             col.names = c("Metric", "na.rm = F", "na.rm = T"),
             align = "lcc",
             caption = "na.rm = T vs. na.rm = F")
```

## Exercise 3
Recreate the frequency plot of scheduled_dep_time colored by whether the flight was cancelled or not. Also facet by the cancelled variable. Experiment with different values of the scales variable in the faceting function to mitigate the effect of more non-cancelled flights than cancelled flights.
```{r}
#| message: false
#| warning: false
df_fli <- nycflights13::flights

p_fli <- df_fli |> 
  mutate(cancelled = is.na(dep_time),
         sched_hour = sched_dep_time %/% 100,
         sched_min = sched_dep_time %% 100,
         sched_dep_time = sched_hour + (sched_min / 60)) |> 
  ggplot(aes(x = sched_dep_time)) + 
  geom_freqpoly(aes(color = cancelled), binwidth = 1/4) +
  labs(title = "Original plot",
       x = "",
       y = "",
       color = "Canceled?") +
  theme(strip.text = element_text(face = "bold", size = 8),
        title = element_text(size = 8, face = "plain"),
        legend.title = element_text(face = "plain", size = 6))

pr_fli <- df_fli |>
  mutate(cancelled = is.na(dep_time),
         sched_hour = sched_dep_time %/% 100,
         sched_min = sched_dep_time %% 100,
         sched_dep_time = sched_hour + (sched_min / 60)) |> 
  ggplot(aes(x = sched_dep_time)) + 
  geom_freqpoly(aes(color = cancelled), binwidth = 1/4) +
  facet_wrap(~ cancelled, scales = "free_y") +
  labs(title = "Recreated plot",
       x = "",
       y = "",
       color = "Canceled?") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 8),
        title = element_text(size = 8, face = "plain"),
        legend.title = element_text(face = "plain", size = 6))

grid.arrange(p_fli, pr_fli, nrow = 2,
             top = grid::textGrob("Fligths frequency by schedule time",
                                   gp = grid::gpar(fontsize = 12, fontface = "bold")),
             left = grid::textGrob("Fligth count",
                                   rot = 90,
                                   gp = grid::gpar(fontsize = 10, fontface = "bold")),
             bottom = grid::textGrob("Departure schedule time",
                                   gp = grid::gpar(fontsize = 10, fontface = "bold")))
```

# 10.5.1.1 Exercises
## Exercise 1
Use what you’ve learned to improve the visualization of the departure times of cancelled vs. non-cancelled flights.
```{r}
#| message: false
#| warning: false
df_fli |>
  mutate(cancelled = is.na(dep_time),
         sched_hour = sched_dep_time %/% 100,
         sched_min = sched_dep_time %% 100,
         sched_dep_time = sched_hour + (sched_min / 60)) |> 
  ggplot(aes(x = sched_dep_time)) + 
  geom_freqpoly(aes(color = cancelled), binwidth = 1/4) +
  facet_wrap(~ cancelled, scales = "free_y") +
  labs(title = "Fligths frequency by schedule time",
       x = "Departure schedule time",
       y = "Fligth count",
       color = "Canceled?") +
  theme_light() +
  theme(legend.position = "top",
        strip.text = element_blank(),
        title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 10, face = "bold"),
        legend.title = element_text(face = "plain", size = 8))
```

## Exercise 2
Based on EDA, what variable in the diamonds dataset appears to be most important for predicting the price of a diamond?  
```{r}
#| message: false
#| warning: false
## Loop
# 1. Slecionando as variáveis
df_rn <- df |> 
  rename("Total Depth" = depth, "Table" = table, "Carat" = carat, "Length" = x, "Width" = y, "Depth" = z)

y_vars <- df_rn |> 
  select(where(is.numeric)) |> 
  names() |> 
  setdiff("price")

# 2. Criar uma lista vazia para guardar os gráficos
plot_list <- list()

# 3. Rodar o loop
for (y_var in y_vars) {
  plot_list[[y_var]] <- ggplot(df_rn, aes(x = price, y = .data[[y_var]])) + 
    geom_jitter(color = "gray20", size = 0.75, alpha = 0.1) + 
    geom_smooth(method = "auto", se = TRUE, color = "#cb73fa") +
    labs(y = y_var) +
    theme_classic() + 
    theme(axis.title = element_text(size = 8, face = "bold"),
          axis.title.x = element_blank())
}

# 4. Grid com todos os gráfico
do.call(grid.arrange, c(plot_list, 
                        list(ncol = 3, 
                             top = grid::textGrob("Diamond price VS other variables", 
                                                  gp = grid::gpar(fontsize = 12, fontface = "bold")),
                             bottom = grid::textGrob("Price",
                                            gp = grid::gpar(fontsize = 10, fontface = "bold")))))
```

## Exercise 3
How is that variable correlated with cut?
```{r}
#| message: false
#| warning: false
df |>  ggplot(aes(x = price, y = carat, color = cut)) +
  geom_jitter(size = 1.2, alpha = 0.7) +
  scale_color_brewer(palette = "PuOr") +
  geom_smooth(method = "auto", se = TRUE, color = "black") +
  theme_minimal() +
  labs(y = "Price",
       x = "Carat",
       color = "Cut",
       title = "Price vs. Carat by Cut") +
  theme(title = element_text(face = "bold", size = 10),
        plot.title = element_text(size = 12))
```

## Exercise 4
Why does the combination of those two relationships lead to lower quality diamonds being more expensive?
```{r}
# Prices vs. the discrete variables
## Cut
cut <- df |> ggplot(aes(x = cut, y = price, fill = cut)) +
  geom_boxplot(color = "black", alpha = 0.7, show.legend = FALSE) +
  scale_y_continuous(breaks = seq(0, 25000, 2500)) +
  scale_fill_brewer(palette = "BuPu") +
  theme(legend.position = "none") +
  labs(x = "Cut",
       y = "Price",
       title = "Price by cut distribuition") +
  theme_minimal()

## Clarity
clrt <- df |> ggplot(aes(x = clarity, y = price, fill = clarity)) +
  geom_boxplot(color = "black", alpha = 0.7, show.legend = FALSE) +
  scale_fill_brewer(palette = "BuPu") +
  scale_y_continuous(breaks = seq(0, 25000, 2500)) +
  labs(x = "Clarity",
       y = "Price",
       title = "Price by clarity distribuition") +
  theme_minimal()

## Color
clr <- df |> ggplot(aes(x = color, y = price, fill = color)) +
  geom_boxplot(color = "black", alpha = 0.7, show.legend = FALSE) +
  scale_fill_brewer(palette = "BuPu") +
  scale_y_continuous(breaks = seq(0, 25000, 2500)) +
  labs(x = "Color",
       y = "Price",
       title = "Price by color") +
  theme_minimal()

crct <- df |> ggplot(aes(x = cut, y = carat, fill = cut)) +
  geom_boxplot(color = "black", alpha = 0.7, show.legend = FALSE) +
  scale_fill_brewer(palette = "BuPu") +
  theme(legend.position = "none") +
  labs(x = "Cut",
       y = "Carat",
       title = "Carat by cut distribuition") +
  theme_minimal()

grid.arrange(crct, cut, clrt, clr, 
             ncol = 2)
```

## Exercise 5
Instead of exchanging the x and y variables, add coord_flip() as a new layer to the vertical boxplot to create a horizontal one. How does this compare to exchanging the variables?
```{r}
# Changing variables
change <- df |> ggplot(aes(x = carat, y = cut, fill = cut)) +
  geom_boxplot(color = "black", alpha = 0.7, show.legend = FALSE) +
  scale_fill_brewer(palette = "BuPu") +
  theme(legend.position = "none") +
  labs(x = "Carat",
       y = "Cut",
       title = "Exchanging") +
  theme_bw()

# Using coord_flip
flip <- df |> ggplot(aes(x = cut, y = carat, fill = cut)) +
  geom_boxplot(color = "black", alpha = 0.7, show.legend = FALSE) +
  scale_fill_brewer(palette = "BuPu") +
  theme(legend.position = "none") +
  labs(x = "Cut",
       y = "Carat",
       title = "coord_flip") +
  theme_bw() +
  coord_flip()

grid.arrange(change, flip, ncol = 2)
```

## Exercise 6
One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”. One approach to remedy this problem is the letter value plot. Install the lvplot package, and try using geom_lv() to display the distribution of price vs. cut. What do you learn? How do you interpret the plots?
```{r}
#install.packages("lvplot")
library(lvplot)

df |> ggplot(aes(x = reorder(cut, price, FUN = median), y = price, fill = cut)) +
  geom_lv(color = "black", alpha = 0.7) +
  scale_fill_brewer(palette = "BuPu") +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "Price vs Cut: Letter-Value Plot",
       x = "Cut",
       y = "Price")
```

## Exercise 7
Create a visualization of diamond prices vs. a categorical variable from the diamonds dataset using
### geom_violin()
```{r}
#| message: false
#| warning: false
# geom_violin()
df |> ggplot(aes(x = cut, y = price, fill = cut)) +
  geom_violin(alpha = 0.7, show.legend = FALSE) +
  scale_fill_brewer(palette = "BuPu") +
  theme_minimal() +
  labs(title = "Violin Plot - Cut vs. Price",
       x = "Cut",
       y = "Price")
```

### geom_histogram()
```{r}
#| message: false
#| warning: false
# geom_histogram()
df |> ggplot(aes(x = price, fill = cut)) +
  geom_histogram(bins = 30, show.legend = FALSE) +
  facet_wrap(~cut, scales = "free_y") +
  scale_fill_brewer(palette = "BuPu") +
  theme_minimal() +
  labs(title = "Faceted Histogram - Price distribuition",
       x = "Price",
       y = "Count")
```

### geom_freqpoly()
```{r}
#| message: false
#| warning: false
# geom_freqpoly()
df |>  ggplot(aes(x = price, color = cut)) +
  geom_freqpoly(binwidth = 500, linewidth = 1) +
  scale_color_brewer(palette = "BuPu") +
  theme_minimal() +
  labs(title = "Frequency Polygon - Price frequency",
       x = "Price",
       y = "Count")
```
### colored geom_density()
```{r}
#| message: false
#| warning: false
# geom_density()
df |>  ggplot(aes(x = price, fill = cut)) +
  geom_density(alpha = 0.4) +
  scale_fill_brewer(palette = "BuPu") +
  theme_minimal() +
  labs(title = "Density Plot - Price Density",
       x = "Price",
       y = "Density")
```

### Compare and contrast the four plots. What are the pros and cons of each method of visualizing the distribution of a numerical variable based on the levels of a categorical variable?

## Exercise 8
If you have a small dataset, it’s sometimes useful to use geom_jitter() to avoid overplotting to more easily see the relationship between a continuous and categorical variable. The ggbeeswarm package provides a number of methods similar to geom_jitter(). List them and briefly describe what each one does.
```{r}
# Creating a small data set for comparation
set.seed(123)

df.tiny <- diamonds |> sample_n(size = 50, replace = F)

# Geom_jitter vs. beeswarm
p1 <- df.tiny |> ggplot(aes(x = color, y = price, color = color)) +
  geom_jitter(size = 2, alpha = 0.5) +
  labs(title = "Jitter") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title = element_blank())

p2 <- df.tiny |> ggplot(aes(x = color, y = price, color = color)) +
  geom_beeswarm(size = 2, alpha = 0.5) +
  labs(title = "Beeswarm",
       color = "Color") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        legend.position = "none")

grid.arrange(p1, p2,
             ncol = 2,
             top = grid::textGrob("Diamond price (in USD) vs. Color",
                                  gp = grid::gpar(fontsize = 12, fontface = "bold")),
             left = grid::textGrob("Price",
                                   rot = 90,
                                   gp = grid::gpar(fontsize = 10, fontface = "bold")),
             bottom = grid::textGrob("Color",
                                     gp = grid::gpar(fontsize = 10, fontface = "bold")))
```

# 10.5.2.1 Exercises
## Exercise 1
How could you rescale the count dataset above to more clearly show the distribution of cut within color, or color within cut?

## Exercise 2
What different data insights do you get with a segmented bar chart if color is mapped to the x aesthetic and cut is mapped to the fill aesthetic? Calculate the counts that fall into each of the segments.

## Exercise 3
Use geom_tile() together with dplyr to explore how average flight departure delays vary by destination and month of year. What makes the plot difficult to read? How could you improve it?

# 10.5.3.1 Exercises
## Exercise 1
Instead of summarizing the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using cut_width() vs. cut_number()? How does that impact a visualization of the 2d distribution of carat and price?

## Exercise 2
Visualize the distribution of carat, partitioned by price.

## Exercise 3
How does the price distribution of very large diamonds compare to small diamonds? Is it as you expect, or does it surprise you?

## Exercise 4
Combine two of the techniques you’ve learned to visualize the combined distribution of cut, carat, and price.

## Exercise 5
Two dimensional plots reveal outliers that are not visible in one dimensional plots. For example, some points in the following plot have an unusual combination of x and y values, which makes the points outliers even though their x and y values appear normal when examined separately. Why is a scatterplot a better display than a binned plot for this case?